### 2022-04-16 paper/Chapters/Conclusion.tex
Feedback from advisor and thesis defense

Should this project be continued in the future, there are multiple ways that additional time and resources could further improve this project.
The first improvement to make is to fix the issue with USB data described in section \ref{subsec:Manufacturing3}.
This is the only flaw in the hardware preventing it from being considered a completed product.
Similar to the previous section, if more time were available, more testing could be done in various environments to identify any weaknesses in the software that may have gone unnoticed.
After experimenting with streaming over different networks, it would be beneficial to look at polishing the hardware of the project into a self-contained product.
As it stands, it is currently a PCB with peripherals attached that make it cumbersome to carry around.
The groundwork for this has already been laid out in Section \ref{subsec:DesigningHDMI}, with FFC connectors established for one HDMI and one USB port for connection to an embedded screen, and with the footprint of the final PCB and battery fitting within the footprint of the chosen screen.
All that is left is to build an enclosure to house the final product.

>>>>

Should this project be continued in the future, there are multiple ways that additional time and resources could further improve this project.
The first improvement to make is to fix the issue with USB data described in section \ref{subsec:Manufacturing3}.
This is the only flaw in the hardware preventing it from being considered a completed product.
Similar to the previous section, if more time were available, more testing could be done in various environments to identify any weaknesses in the software that may have gone unnoticed.

Another aspect of the project worth testing in more detail is the performance of the hardware produced in Chapter \ref{Chapter4} compared to the hardware performance of the development board that is sold with the CM4 (Section \ref{sec:ChoosingParts}).
The ideal outcome is that the two boards provide the same performance, meaning that there is no unneeded overhead introduced by the board produced in this thesis that should be fixed.
In the event that there is slowdown somewhere within the custom board, the specific problem should be identified and redesigned to provide optimal performance.

After experimenting with streaming over different networks and testing the project under different conditions, it would be beneficial to look at polishing the hardware of the project into a self-contained product.
As it stands, it is currently a PCB with peripherals attached that make it cumbersome to carry around.
The groundwork for this has already been laid out in Section \ref{subsec:DesigningHDMI}, with FFC connectors established for one HDMI and one USB port for connection to an embedded screen, and with the footprint of the final PCB and battery fitting within the footprint of the chosen screen.
All that is left is to build an enclosure to house the final product.

### 2022-04-16 paper/Chapters/Evaluation.tex

\todoadvisor{How many samples were taken to create the figures? How many times was the test run?}

@@

\begin{enumerate}
  \item Both the host and client computers are fully rebooted to ensure no other application is running.
  \item The host computer starts up a single browser window with the web page described in Section \ref{sec:DevelopingTestingAndMeasurementTools}.
  \item If the test calls for the CPU or the GPU to be stressed, the application \enquote{Blender} is started on the host machine, and a benchmark is performed for the CPU or the GPU.
        This benchmark will run for longer than the duration of the test, and will keep the CPU or GPU at 100\% load for the entire duration of the benchmark.
  \item Both devices start up their respective data collection tool that will record key-strokes and changes in the screen's color.
  \item The client machine connects to the host computer using the application being tested.
  \item The user presses the \enquote{0} key on the client's numpad to synchronize \todoadvisor{Syncronize what?}the two devices' data recording tools.\todoadvisor{How far off are they?}
  \item The user begins by typing on the client's keyboard \todoadvisor{what does slowly mean?}slowly, allowing both computers to record their data.
  \item The user then types on the client's keyboard quickly to ensure that \todoadvisor{Why is there a seperation here?}quick successive inputs are handled correctly and no data is lost or duplicated.
  \item Once typing is complete, both the client and the host computers close their data recording tools and the client ends the streaming session.
\end{enumerate}

@@

\noindent
Once all the data is collected, the data from the host and client machines for a single trial are organized into a single CSV file for processing.
In order to ensure that the data is comparable, all extraneous factors must be removed from the data.
Namely, the network latency and difference in computer clocks must be removed.
This is partially accounted for in step \emph{6} of the above process, and the remaining inaccuracy can now be removed by subtracting the time difference between the when the client presses the button and the time the host executes that command.
This ensures that any inaccuracies in each computer's internal clock will not impact the data.

@@

The first factor in determining the performance of the application is to take a look at how responsive it is to use.
Responsivity is defined as how quick the application responds to inputs and how quickly the user can see the results of their inputs.
This is important for general user experience and applications where continuous input is required.
This is also what is generally referred to when determining whether an application has a positive user experience, since a program that can take seconds to respond to the user begets lower productivity and a more frustrating experience.
To analyze this, two points of data are recorded: First, the amount of time between when the client records a keystroke and when it is actually inputted on the host machine, and secondly, the amount of time between the host machine updating their screen and when the client actually displays that update.
The first datum is generally referred to as the latency of the application, since even though each trial is done using the same network setup different programs may have different optimizations built in that reduce the time needed for the host to implement keystrokes made by the client.
Figure \ref{fig:InputDelay} shows the amount of time between the client and the host registering the keystroke under the different test conditions.

@@

      \legend{Ideal,CPU stressed,GPU stressed}
    \end{axis}
  \end{tikzpicture}
  \caption[Color Delay Data]{Time for client to render update to the screen.}
  \label{fig:ColorDelay}
\end{figure}
\todoadvisor{Maybe put full figure in the appendix?}

>>>>

@@

\begin{enumerate}
  \item Both the host and client computers are fully rebooted to ensure no other application is running.
  \item The host computer starts up a single browser window with the web page described in Section \ref{sec:DevelopingTestingAndMeasurementTools}.
  \item If the test calls for the CPU or the GPU to be stressed, the application \enquote{Blender} is started on the host machine, and a benchmark is performed for the CPU or the GPU.
        This benchmark will run for longer than the duration of the test, and will keep the CPU or GPU at 100\% load for the entire duration of the benchmark.
  \item Both devices start up their respective data collection tool that will record key-strokes and changes in the screen's color.
  \item The client machine connects to the host computer using the application being tested.
  \item The tester presses the \enquote{0} key on the client's numpad to synchronize the two devices' clocks and data recording tools (Detailed below).
  \item The tester begins by typing on the client's keyboard at a steady pace as close to 50 words per minute as possible, allowing both computers to record their data.
  \item The tester then types on the client's keyboard at a quicker pace of approximately 100 words per minute to ensure that quick successive inputs are handled correctly and no data is lost or duplicated.
  Due to the wide variance in what the user can do using the software, it is nearly impossible to test real world scenarios.
  Instead, high frequency inputs coupled with updating the entire screen mimic high-intensity applications.
  \item Once typing is complete, both the client and the host computers close their data recording tools and the client ends the streaming session.
\end{enumerate}

@@

\noindent
Once all the data is collected, the data from the host and client machines for a single trial are organized into a single CSV file for processing.
In order to ensure that the data is comparable, all extraneous factors must be removed from the data.
Namely, the network latency and difference in computer clocks must be removed.
By keeping the host and client computers attached to the same network through ethernet, the network latency is minimized.
But the task of keeping two computer's internal clocks synchronized is not trivial.
Even after instructing the operating system to synchronize its clock with an internet time server, the two computers may still report different times.
In some cases, the difference between the host and client computer's system time was noticed to be upwards of a full second after resynchronizing both clocks \cite{time.is}.
This is partially accounted for in step \emph{6} of the above process, and the remaining inaccuracy can now be removed by subtracting the time difference between the when the client presses the button and the time the host executes that command.
This ensures that any inaccuracies in each computer's internal clock will not impact the data.

@@

The first factor in determining the performance of the application is to take a look at how responsive it is to use.
Responsivity is defined as how quick the application responds to inputs and how quickly the user can see the results of their inputs.
This is important for general user experience and applications where continuous input is required.
This is also what is generally referred to when determining whether an application has a positive user experience, since a program that can take seconds to respond to the user begets lower productivity and a more frustrating experience.
To analyze this, two points of data are recorded: First, the amount of time between when the client records a keystroke and when it is actually inputted on the host machine, and secondly, the amount of time between the host machine updating their screen and when the client actually displays that update.
The first datum is generally referred to as the latency of the application, since even though each trial is done using the same network setup different programs may have different optimizations built in that reduce the time needed for the host to implement keystrokes made by the client.
Each of the following figures shows data over one trial for each category.
Each trial lasted approximately 60 seconds following the steps detailed in Section \ref{sec:TestingMethodology}, totaling to approximately 200 keystrokes per trial.
Limitations in this data is discussed in Section \ref{sec:ConclusionLimitations}.
Figure \ref{fig:InputDelay} shows the amount of time between the client and the host registering the keystroke under the different test conditions.

@@

      \legend{Ideal,CPU stressed,GPU stressed}
    \end{axis}
  \end{tikzpicture}
  \caption[Color Delay Data]{Time for client to render update to the screen.}
  \label{fig:ColorDelay}
\end{figure}

### 2022-04-16 paper/Chapters/Software.tex

The second half of the project is composed of software that is used to connect the client machine to a host computer.
Designing the software begins with identifying it's requirements in Section \ref{sec:SoftwareRequirements}.
Then, potential avenues for development are discussed in Section \ref{sec:PotentialAvenues} so that the software does not need to be built from scratch.
Following that, special considerations must be made while developing the software for the CM4 running on an \emph{ARM} processor in Section \ref{sec:DevelopingForARM}, and the development of the suite of tools used to record data is detailed in Section \ref{sec:DevelopingTestingAndMeasurementTools}.
\todo{Link git repo}

@@

\section{Potential Avenues}\label{sec:PotentialAvenues}

As with any software project, there are many potential ways to solve the problem.
In many cases, the easiest place to start is to begin by ruling out potential avenues that are not feasible.
For example, as evident by the number of existing solutions that attempt to solve this problem (Chapter \ref{Chapter3}), taking into special account Nvidia's recent foray into remote gaming with the Shield (\ref{subsec:NvidiaShieldAndGameStream}), it is not reasonable to attempt to create a brand new software solution from scratch with the time and resources available.
It has taken Nvidia, a company worth hundreds of billions of dollars, years to get to the point they are at now \cite{brown_2013}.
Developing a rival protocol or new software platform is simply not possible for this project.
Instead, given that Nvidia's GameStream technology is the current leader in the realm of high-performance remote computing, the best way to build a software product that meets the requirements laid out in Section \ref{sec:ResearchQuestions} is to build upon what already exists.

\subsection{NVIDIA GameStream}\label{subsec:NVIDIAGameStream}

\todo[linecolor=green,backgroundcolor=green!25,bordercolor=green,inline]{Should this subsection and the following subsection be top level sections? The organization of the headers is a little strange, but the content flows pretty well}

@@

\subsection{Moonlight}\label{subsec:Moonlight}

\todoadvisor{Mention this earlier, maybe to the state of the art}
Moonlight is an unofficial third-party open-source client for the Nvidia GameStream protocol \cite{moonlight}.
The project is structured as a core library written in \emph{C} with a number of community built clients built on top of it for various platforms.
In essence, it turns the process described in the previous section (\ref{subsec:NVIDIAGameStream}) into a usable programming interface.
By reverse engineering the protocol originally implemented by the Nvidia Shield, Moonlight enables the development of a GameStream client for almost any operating system.
The Moonlight community has already developed fantastic clients for many platforms such as Windows, Mac, Linux, Android, iOS, Web, and even other gaming consoles.
All that is needed for the purposes of this project is to make a fork of the project that tweaks it to best suit the CM4 that drives the hardware and develop a suite of tools to test it's functionality.

@@

In order to create an out-of-the-box experience for someone using the hardware for this project, a custom operating system image is created that has all the dependencies, libraries, and applications pre-installed and pre-configured.
Manjaro provides good tools for building and managing custom images of the Manjaro operating system for \emph{ARM} architectures \cite{manjaro_arm_tools}, with the only drawback being that it must be run on an \emph{ARM} processor, and is much more difficult to run on another architecture.
To make development easier, a toolchain was built that could be run on the CM4 to build the code that is written on other computers.
All the developer needs to do is to pull the git repository\todoadvisor{Cite the git repository} of this project onto the \emph{ARM} architecture, run a set up script to install dependencies, and then run the build script to compile everything into an immediately deployable OS image.
This build script will automatically compile all the necessary code using \emph{cmake}, package all the necessary files using \emph{makepkg}, and build a fresh OS image that is ready to flash.
The following excerpt details the three functions used to make this work.

@@

\todoadvisor{Explicitly mention that this is work I did}
In order to evaluate the performance of the project, a suite of tools must be developed to test both the newly developed application and the existing solutions.
To do this, two sets of data will be collected: First, data is recorded directly from the forked version of moonlight that logs statistics such as the number of frames received, dropped, or rendered, and data that is recorded from a separate application that logs the amount of time it takes for the client machine to receive updates after it sends data to the host machine.
The first set of data is important to track the raw performance of the application, with statistics like Frames Per Second (FPS) and decoding time which are difficult to record otherwise.
These statistics provide insight into how performant the application is from a technical perspective.
The following is a structure that holds the statistics recorded over a session of connecting to a host computer.
The data recorded is discussed in Chapter \ref{Chapter6}.

@@

The second set of data is recorded by measuring the total amount of time it takes for the host to receive data from the client, and for the client to receive updates back from the host.
This is done using the following steps once the client is connected to the host:
\todoadvisor{Mention that the times need to be syncronized}

@@

This records two points of data, the time it takes for the host to implement the actions done by the user, and the time that the user sees the result of their actions.
Because this method of data collection runs as a separate application on both the client and the server, it can be used to measure the performance of any of the remote desktop applications for comparison.
This is again discussed in Chapter \ref{Chapter6}.

@@

This is sufficient to change the color of the screen back and forth between easily distinguishable colors for both the client and the host to record.
\todoadvisor{Mention that this is recorded by the dataloggerz}
With both of these datasets recorded, the next step is to analyze the data and evaluate the performance of the application.

>>>>

The second half of the project is composed of software that is used to connect the client machine to a host computer.
Designing the software begins with identifying it's requirements in Section \ref{sec:SoftwareRequirements}.
Then, the Nvidia GameStream Protocol is analyzed in depth in Section \ref{sec:UtilizingNVIDIAGameStream} so that the software does not need to be built from scratch.
Following that, special considerations must be made while developing the software for the CM4 running on an \emph{ARM} processor in Section \ref{sec:DevelopingForARM}, and the development of the suite of tools used to record data is detailed in Section \ref{sec:DevelopingTestingAndMeasurementTools}.

The entirety of the code developed for this project is available on GitHub at: \url{https://github.com/Dylancyclone/thesis/code}.

@@

\section{Utilizing NVIDIA GameStream}\label{sec:UtilizingNVIDIAGameStream}

As with any software project, there are many potential ways to solve the problem.
As evident by the number of existing solutions that attempt to solve this problem (Chapter \ref{Chapter3}), taking into special account Nvidia's recent foray into remote gaming with the Shield (\ref{subsec:NvidiaShieldAndGameStream}), it is not reasonable to attempt to create a brand new software solution from scratch with the time and resources available for this thesis.
It has taken Nvidia, a company worth hundreds of billions of dollars, years to get to the point they are at now \cite{brown_2013}.
Developing a rival protocol or new software platform is simply not possible for this project.
Instead, given that Nvidia's GameStream technology is the current leader in the realm of high-performance remote computing, the best way to build a software product that meets the requirements laid out in Section \ref{sec:ResearchQuestions} is to build upon what already exists.

@@

By taking advantage of the Moonlight library briefly described in Section \ref{subsec:NvidiaShieldAndGameStream}, the previously described process is translated  into a usable programming interface.
To accomplish this, a fork of the Moonlight project is created so the GameStream protocol can be leveraged in this thesis.
With the fork, much of the code required to create the software solution is already provided, all that is left is to tweak the code to ease development for the \emph{ARM} architecture and to build a suite of testing tools that hook into Moonlight to extract statistics.

@@

In order to create an out-of-the-box experience for someone using the hardware for this project, a custom operating system image is created that has all the dependencies, libraries, and applications pre-installed and pre-configured.
Manjaro provides good tools for building and managing custom images of the Manjaro operating system for \emph{ARM} architectures \cite{manjaro_arm_tools}, with the only drawback being that it must be run on an \emph{ARM} processor, and is much more difficult to run on another architecture.
To make development easier, a toolchain was built that could be run on the CM4 to build the code that is written on other computers.
All the developer needs to do is to pull the git repository of this project onto a machine running the \emph{ARM} architecture \cite{ProjectGitHub}, run a set up script to install dependencies, and then run the build script to compile everything into an immediately deployable OS image.
This build script will automatically compile all the necessary code using \emph{cmake}, package all the necessary files using \emph{makepkg}, and build a fresh OS image that is ready to flash.
The following excerpt details the three functions used to make this work.

@@

In order to evaluate the performance of the project, a suite of tools must be developed to test both the newly developed application and the existing solutions.
To do this, two sets of data will be collected: First, data is recorded directly from the forked version of moonlight that logs statistics such as the number of frames received, dropped, or rendered, and data that is recorded from a separate application that logs the amount of time it takes for the client machine to receive updates after it sends data to the host machine.
The first set of data is important to track the raw performance of the application, with statistics like Frames Per Second (FPS) and decoding time which are difficult to record otherwise.
These statistics provide insight into how performant the application is from a technical perspective.
The following is a structure that was developed for this thesis to hold the statistics recorded over a session of connecting to a host computer.
Corresponding code was written throughout the fork of the Moonlight application to populate this structure while a streaming session is running.
The data recorded is discussed in Chapter \ref{Chapter6}.

@@

The second set of data is recorded by measuring the total amount of time it takes for the host to receive data from the client, and for the client to receive updates back from the host.
This is done using the following steps once the client is connected to the host:

@@

This records two points of data, the time it takes for the host to implement the actions done by the user, and the time that the user sees the result of their actions.
Because this method of data collection runs as a separate application on both the client and the server, it can be used to measure the performance of any of the remote desktop applications for comparison.
An important obstacle to using this data is the difficulty in synchronizing the host and client machines' clocks.
Even if each machine accurately recorded the amount of time between each event and the order in which they occur, the data is useless if the two machines do not agree at which time the test began.
This is again discussed in Chapter \ref{Chapter6}, with the issue of clock synchronization being discussed in Section \ref{sec:TestingMethodology}.

@@

This is sufficient to change the color of the screen back and forth between easily distinguishable colors for the data loggers of both the client and the host to record.
With both of these datasets recorded, the next step is to analyze the data and evaluate the performance of the application.
